\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{palatino}
\usepackage{helvet}
\usepackage{times}
\usepackage{layout}
\usepackage[a4paper,top=2.0cm, right=2.0cm, bottom=2.0cm, left=2.0cm]{geometry}
\usepackage{enumitem}
\usepackage{amsthm}
\usepackage{url}
\usepackage{multicol,caption}

\setlist{nolistsep}

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{{\hvnb Colm}}
\chead{}
\rhead{}
\lfoot{}
\cfoot{}
\rfoot{}

\newcommand*{\helvetica}{\fontfamily{phv}\selectfont}
\newcommand*{\helveticanarrow}{\fontfamily{phv}\fontseries{mc}\selectfont}
\newcommand*{\hvnb}{\fontfamily{phv}\fontseries{bc}\selectfont}
\newcommand*{\palatino}{\fontfamily{ppl}\selectfont}
\newcommand*{\timesroman}{\fontfamily{ptm}\selectfont}

% Commands to produce formatted layout
\newcommand*{\projecttitle}[1]{\begin{center}\Large\hvnb{\color{blue} #1}\end{center}}
\newcommand*{\theauthor}[1]{\noindent  \helveticanarrow{#1}\\}



% Support for multiple bibliographies
\usepackage[sectionbib,numbers]{natbib}
\usepackage{chapterbib}

\usepackage[compact]{titlesec}
\usepackage{lipsum}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amsfonts}

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

\titleformat{\section}
  {\large\hvnb\color{blue}}{\thesection}{1em}{}
\titleformat{\subsection}
  {\hvnb}{\thesubsection}{1em}{}
\titleformat{\chapter}
  {\Large\hvnb\color{blue}}{Lecture \thechapter:}{1em}{}

\newenvironment{Figure} 
  {\par\medskip\noindent\minipage{\linewidth}}
  {\endminipage\par\medskip}

\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\dd}[2]{\frac{d {#1}}{d {#2}}}
\DeclareMathOperator{\sgn}{sgn}

\begin{document}
\setlength{\bibsep}{0.2pt}
\setlength{\itemsep}{0.2pt}

\helveticanarrow
\rhead{\hvnb  16/06/2026 working notes}

\projecttitle{Notes on ergodicity}

%\theauthor{Colm Connaughton} 


\begin{multicols}{2}

\section{Introduction: the concept of ergodicity}

Does observing a single instance of a system for a long enough time allow one to understand the statistics of a population of such systems?
We say a system is ergodic when the answer to this question is yes.
Ergodicity justifies generalising temporal observations of an individual experiment, agent or organism to population level statistical statements.
It also justifies specialising population level statistical properties to statements about individual members of a population.
The scientific study of ergodicity grew out of physics where it provided justification for the use of ensembles in statistical mechanics and out of statistics where it provided justification for the efficacy of statistical inference.
The reason researchers needed a word for this property is that it was quickly understood by physicists, statisticians and mathematicians that it does not always hold and should therefore not be treated as an unexamined assumption.

Despite the fundamental and fairly intuitive nature of the concept, it turns out to be difficult to make the notion of ergodicity mathematically precise in a way that applies to all the use cases.
This note is my own attempt to piece together the various facets of ergodicity

\section{Mathematical formulation of ergodicity for deterministic dynamical systems}

\begin{itemize}
\item Ergodicity is a property of a measure-preserving dynamical system $(\Omega, \mu, T)$ where
\begin{itemize}
\item $\Omega$ is the phase space - the set of all possible states the system can be in.
\item $\mu$ is a measure which assigns a non-negative volume to any subset of the phase space
\item $T \, :\, \Omega \mapsto \Omega$ is a map which generates dynamics in $\Omega$ by iteration:
\begin{align}
x_{n+1} &= T(x_n)
\end{align}
$T$ is required to be measure-preserving in the sense that
\begin{align}
\mu(T^{-1} A) & = \mu(A)
\end{align}
for all subsets $A \subset \Omega$. 
It is also required to be finite in the sense that $\mu(\Omega) < \infty$. Typically we normalise so that $\mu(\Omega) = 1$.
In many examples, the phase space is compact which makes it easier for the total measure to be finite but this is not a requirement.
\end{itemize}
\item 
The definition of ergodicity which has proven most fruitful mathematically is surprisingly abstract: a measure preserving dynamical system is ergodic if the only invariant sets have measure zero or 1.
Here invariant means $T(A) = A$.
\item
A consequence of this definition is the more familiar statement about equality of time averages and phase space averages as expressed by Birkhoff's Theorem:
\begin{align}
\lim_{N \to \infty} \frac{1}{N} \sum_{n=0}^{N-1} f(T^n x_0) = \int_\Omega f(x) d\mu(x)
\end{align}
for all integrable functions $f$ on $\Omega$.
\item
This can also be thought of as a statement about recurrences (how often a trajectory revisits a set $A \subset \Omega$) : if you follow a single trajectory of the dynamical system and count how often it is in a set $A$, the proportion of iterations the trajectory is in $A$ will converge to  $\mu(A)$. 
Mathematically,  this follows from applying Birkhoff's theorem to the indicator function on $A$,  $1_A(x)$:
\begin{align}
\label{eq-recurrences}
\lim_{N\to \infty} \frac{1}{N} \sum_{n=0}^{N-1} 1_A(T^n x_0) & = \int_\Omega 1_A(x) d\mu = \mu(A).
\end{align}
Recurrence frequencies become stable as $N$ grows and are consistent with the invariant measure. 
Let us denote by $\tau_A$ the random variable corresponding to the time that a trajectory takes to revisit $A$ having left it.
This is called the return time. 
One can show that the mean return time, $\mathbb{E}\left[\tau_A\right] = \frac{1}{\mu(A)}$.
Another way to state Eq.~(\ref{eq-recurrences}) is that for almost all sets, $A$, the number of recurrent visits to $A$ asymptotically grows proportional to $N$ : 
\begin{align}
 \sum_{n=0}^{N-1} 1_A(T^n x_0)  \sim c\, N
\end{align}
where $c>0$ is the inverse of the mean return time.
\item
Ergodicity provides a precise notion of typicality: in an ergodic system almost all initial conditions will eventually explore the full phase space. ``Almost all'' means those atypical trajectories that don't explore the full phase space have measure zero. 
The Goldstein and Lebowitz typicality account of ergodicity asserts that for a broad class of many body physical systems, typical trajectories correspond to thermomodynamic equilibrium states. 
This point of view holds that typicality of equilibrium states is more important for the foundations of statistical mechanics than proving that the underlying equations of motion are ergodic with respect  to the Gibbs measure in the strict mathematical sense.
\item
Ergodicity is a property of the map, $T$, and the invariant measure, $\mu$, but the latter need not be unique.  Dynamical systems research tends to give primacy to the map, $T$, and then ask which invariant measures it admits and whether they are ergodic.  
Physically, different invariant measures correspond to different notions of typicality.
\end{itemize}

\section{Stationary stochastic processes as dynamical systems}

As a physicist, one of the more difficult aspects of ergodic theory is understanding how the above notion of ergodicity developed for deterministic dynamical systems applies to stochastic processes where we tend to think of ergodicity in terms of the equality of time averages and ensemble averages.
In the case of deterministic dynamics, randomness can be introduced only via random choice of initial conditions.
We can then think of the measure $\mu$ (suitably normalised) as a probability measure on the phase space and we have a random dynamical system albeit with the sole source of randomness being the choice of initial condition.
Stochastic processes on the other hand have intrinsically random dynamics for example encoded by the transition kernel of a Markov chain. 
If we naively think of the generator of the a stochastic process as being analogous to the map $T$ in the deterministic case, the definition of ergodicity in terms of invariant sets doesn't seem natural any more since the image of any set is random.
This is the wrong way to think about it.
The conceptual unification of ergodicity in stochastic processes with ergodicity in deterministic dynamical systems happens is via a completely different route: the so-called shift map.

The shift map is a way to construct a deterministic measure-preserving dynamical system from any {\em stationary} stochastic process.
The notion of ergodicity described above for measure preserving dynamical systems applies directly to the shift system and the stochastic processes notion of equivalence of time averages and ensemble averages then follows for appropriate choice of observables.

\begin{itemize}
\item Suppose we have a stationary discrete time stochastic process, $(\ldots, x_{-1}, x_0, x_1, x_2,\ldots)$, where the $x_i$ are elements of an underlying state space, $S$. 
Let us denote the stationary distribution of this process by $\pi_*$.
\item Construct the path space, $\Omega = S^{\mathbb{Z}}$. This is the set of all possible trajectories of the stochastic process. I am deliberately calling this $\Omega$ since it is the 
path space (not the state space, $S$) that plays the role of the phase space in the deterministic case.
\item The law of the stochastic process induces a probability measure, $\mu$, on the path space, $\Omega$ - essentially assigning a probability to each path based on how likely it is.  
If the process is very simple and all $x_i$ are independent then this induced measure is the product measure $\mu = \pi_*^{\mathbb{Z}}$.
For non-trivial cases, however, the measure $\mu$ is potentially very complicated indeed since it encodes all dynamical structure, all temporal correlations and all finite-dimensional marginal distributions.
\item
We now define the shift operator $T\, : \, \Omega \mapsto \Omega$ on the path space. For any trajectory $\omega = (\ldots, x_{-1}, x_0, x_1, x_2,\ldots)$, the $n^\text{th}$ entry of the trajectory $T \omega$ is given by
\begin{align}
\left( T\omega \right) _n & =  \left( \omega \right)  = x_{n+1}.
\end{align}
That is to say, the shift map applied to a trajectory moves the time index forward by one step. 
As a map of the path space, the shift map $T$ is deterministic.
\item
Since the original stochastic process is stationary, $x_{n+1}$ has the same distribution as $x_n$ for all $n$ (invariance of the distribution with respect to time shift is usually taken as the definition of first order stationarity.)
This means that $T$ does not change the measure of individual trajectories or sets of trajectories and is thus measure preserving.
\item The path space, $\Omega$, together with the induced probability measure, $\mu$ and the shift map, $T$, therefore constitutes a measure preserving dynamical system. It is ergodic if the only invariant sets are trivial.
Note that unlike for the deterministic case where lots of different dynamics, $T$, are studied, for stochastic processes, $T$ is always the shift map (although this map obviously depends in a complicated way on the generator of the process.) Similarly, the analogue of the phase space is always the path space not the underlying state space of the stochastic process.
\item
If the shifted system is ergodic then Birkhoff's Theorem holds:
\begin{align}
\lim_{N \to \infty} \frac{1}{N} \sum_{n=0}^{N-1} F(T^n \omega) = \int_\Omega f(\omega) d\mu(\omega)
\end{align}
for any trajectory $\omega \in \Omega$ and any integrable function, $F$ on path space.
\item 
However, we do not usually think of time and ensemble averages over trajectories in path space.
To obtain the more usual expression of Birkhoff's theorem for stochastic processes, we choose particular observables that pick out regular functions of single variables, $F(\omega) = f((\omega)_0)$.
Since this observable depends only on a single coordinate, in this case $(\omega)_0 = x_0$, the integral on the RHS only depends on the one-dimensional marginal of the path space induced measure $\mu$. 
Moreover, since the process is stationary, this one-dimensional marginal distribution is the stationary distribution of the original stochastic process.
We then have
\begin{align}
\lim_{N \to \infty} \frac{1}{N} \sum_{n=0}^{N-1} f(T^n x_0) = \int_S f(x) d \pi_*(x) = \mathbb{E}_{\pi_*}[f].
\end{align}
This is the more usual form of Birkhoff's Theorem as stated for ergodic stochastic processes.
\item
We can again think of this in terms of recurrences: given a set $A \subset S$ (an event) if you observe a single realisation of the stochastic process and count the proportion of time the trajectory visits the set $A$, this will converge to probability of the event $A$ under the stationary distribution of the process.
\item
Give example of discrete analogue of Ornstein-Uhlenbeck process driven by Bernoulli noise:
\begin{align}
x_{n+1} &= a\,x_n + \xi_{n+1}
\end{align} 
where $\xi_n \in  \left\{+1, -1 \right\}$ with equal probability.
For $a = \frac{1}{2}$ for example the stationary density,  $\pi_* = \text{Unif}(-2,2)$.  Time averages and ensemble averages for simple moments can be calculated explicitly and shown to agree.

\end{itemize}

\section{Infinite ergodicity}

Infinite ergodic theory extends some aspects of ergodic theory to recurrent dynamics for which an invariant measure exists but is not normalisable, $\mu(\Omega) = \infty$.
A characteristic feature is that almost all sets, $A \subset \Omega$,  are visited infinitely often but the mean return time, $\mathbb{E}\left[\tau_A\right]$,  is infinite. 
Mathematically, this is happens when the distribution of $\tau_A$ has sufficiently heavy tails that $\mathbb{E}\left[ \tau_A\right]$ does not exist.
Physically it means that for almost all sets, $A$,  the dynamics visits $A$ infinitely many times but nevertheless spends only a vanishing proportion of the {\em total} time in $A$.
This is possible because excursions from $A$ can be very long.

Birkhoff's Theorem is often trivial because time averages and recurrence frequencies tend to zero.
Infinite ergodic theory then has less to do with equality of time and ensemble averages. 
Instead of the time-average of an observable, $f$, the central object of study is the time cumulative sum of $f$:
\begin{align}
\label{eq-timeCumulativeSum}
S_N f(x) &= \sum_{n=0}^{N-1} f(T^n x). 
\end{align}
For finite ergodic systems, time cumulative sums grow linearly with $N$ at a rate which is asymptotically deterministic:
\begin{align}
S_N f(x)  &\sim N \int_\Omega f d\mu.
\end{align}
In infinite ergodic systems, by contrast, time cumulative sums grow sub-linearly with $N$ at a rate which remains random even asymptotically:
\begin{align}
S_N f(x) & \sim a_N \, \zeta
\end{align}
where $a_N < N$ and $\zeta$ is a random variable with a non-trivial distribution.
The sub-linear scaling of time cumulative sums explains why ordinary time averages and recurrence frequencies go to zero in such systems. 
Asymptotic randomness means that different trajectories retain order 1 differences forever.
This feature makes infinite ergodic systems fundamentally different from a physical perspective.

In stochastic processes, the first example of infinite recurrence times that often comes to mind is the simple random walk on $\mathbb{Z}$.
Does the simple random walk provide an illustrative example of infinite ergodicity?
I would argue this is not immediately obvious because the simple random walk is not stationary. 
We cannot proceed as in the previous section by taking the invariant measure to be the probability measure induced on path space by the stationary probability distribution of the stochastic process. 
However we do not require the invariant measure to be a probability measure (normalisable). 
What alternative does this enable?
The transition probabilities for the simple random walk are:
\begin{align}
\label{eq-SRWtransitions}
P(i,j) = \left\{
\begin{array}{ll}
\frac{1}{2}&\text{if $i=j+1$}\\
\frac{1}{2}&\text{if $i=j-1$}\\
0&\text{otherwise.}
\end{array}
\right.
\end{align}
The so-called counting measure on $\mathbb{Z}$ is defined as follows.
For any $A \subset \mathbb{Z}$:
\begin{align}
\mu(A) = \left\{  
\begin{array}{ll}
| A | & \text{if $A$ is finite}\\
\infty & \text{otherwise}
\end{array}
\right.
\end{align}
In particular, $\mu(\left\{ i \right\} )=1$ for all $i \in \mathbb{Z}$.
Note that although the counting measure on $\mathbb{Z}$ is not normalisable, it is invariant with respect to the transition probabilities Eq.~(\ref{eq-SRWtransitions}):
\begin{align*}
\sum_{j \in \mathbb{Z}} P(i, j) \mu(\left\{ j \right\} ) &= \frac{1}{2} \mu(\left\{ i -1 \right\} ) + \frac{1}{2} \mu(\left\{ i+1 \right\} )\\
& = \frac{1}{2} + \frac{1}{2}  = \mu(\left\{ i \right\} ).
\end{align*}
Physically, this simply means that at each step, there is as much probability mass entering each site  $i \in \mathbb{Z}$  from neighbouring sites as there is leaving due to hopping.
The counting measure on $\mathbb{Z}$ can now be used to induce an invariant (but non-normalisable) measure on the path space of the simple random walk. 
A slightly unintuitive aspect of this construction is that this induced measure on path space is stationary even though the original stochastic process is not.
By taking $f(x) = 1_{\left\{0\right\}}(x)$ in Eq.~(\ref{eq-timeCumulativeSum}), we can count the number of visits to site $0$ after time $N$.
From the theory of random walks one finds
\begin{align}
S_N f \sim c \, \sqrt{N} 
\end{align}
where $c$ has a non-trivial (half normal?) distribution.
Thus ordinary frequencies vanish but normalised occupation counts remain meaningful. 

To avoid giving the impression that infinite ergodicity is a complicated way of describing properties of random walks that are obtainable by other means, it is worth briefly considering a second example: the Pomeau-Manneville map. 
This is a deterministic dynamical system with a compact phase space, $\left[0,1\right]$ where infinite invariant measure arises in a very different way to the simple random walk.
The map is
\begin{align}
T(x) &= x + x^{1+\alpha}\ \mod\ 1
\end{align}
where $x \in \left[0,1\right]$ and $\alpha>0$.
The important feature is that the fixed point at $0$ is neutrally stable because $T^\prime(0) = 1$.
Consequently trajectories that wander close to zero can get trapped for very long periods of time.
How long depends on the parameter $\alpha$.
One can show that an invariant measure exists and that near zero it behaves like
\begin{align}
\label{eq-PMrho}
\rho(x) &\sim x^{-\alpha},
\end{align}
which reflects the fact that trajectories have a tendency to accumulate in the vicinity of the ``sticky'' neutral fixed point.
If we consider a set which does not include the neutral fixed point at $0$, for example $A = \left[ \frac{1}{4}, \frac{3}{4}\right]$, it is possible to show that the distribution of the return time, $\tau_A$, has power law tails:
\begin{align}
\mathbb{P}(\tau_A > n) & \sim n^{-\frac{1}{\alpha}}.
\end{align}
These tails are the result of some trajectories getting trapped near the fixed point and taking a long time to escape.
The parameter value $\alpha=1$ is special. 
If $\alpha <1$ the return time distribution is still integrable, the invariant measure is finite and ordinary ergodic theory applies. 
If $\alpha >=1$, the mean return time diverges and we enter the infinite ergodic theory regime.
Note from Eq.~(\ref{eq-PMrho}) that the invariant measure becomes infinite since $\int_0^\epsilon \rho(x)\,dx$ diverges for any $\epsilon > 0$.
In this example, infinite measure arises due to the loss of an integrable singularity at a single point in phase space rather than due to trajectories spreading out over infinitely large regions in the case of the random walk example. 
Nevertheless many of the same phenomena are observed as we noted for the random walk example.
Cumulative occupation counts again scale sub-linearly and after appropriate normalisation converge to a nontrivial random limit.
This is the aspect that generalises across many examples: infinite ergodic theory is fundamentally about how power law recurrence time statistics lead to persistent randomness in a broad class of deterministic and random dynamical systems.

\section{Nonstationary stochastic processes}

Most nonstationary processes are not recurrent so the notion of infinite ergodicity does not apply to them. 
Technically, it is tautological to refer to such processes as non-ergodic since nonstationarity precludes ergodicity in the original sense and non-recurrence precludes the more general notion of infinite ergodicity.
Nevertheless in applied contexts, authors sometimes emphasise the non-ergodicity of some ``obviously non-ergodic''  nonstationary dynamics.
This can be because the question of whether individual realisations are representative of the ensemble remains important. 
It can also be because it is possible to find a transformation which isolates a useful ergodic component of the dynamics, either exactly or approximately.

An illustrative example is provided by the multiplicative random walk:
\begin{align}
\label{eq-GRW}
X_{n+1} &= z_n\,X_n
\end{align}
with $X_0 >0$ and the $z_i$ being iid non-negative random variables with finite mean and variance (the case $z_i \sim \text{Bernoulli}\left[\frac{3}{5}, \frac{3}{2}\right]$ is the multiplicative cointoss example.)
We have
\begin{align}
X_n = X_0 \prod_{i=1}^{n-1} z_i
\end{align}
which is clearly nonstationary and therefore nonergodic.
Let us take the logarithm and divide by $n$:
\begin{align}
\frac{\log{X_n} - \log{X_0}}{n} & = \frac{1}{n} \, \sum_{i=1}^{n-1} \log{z_i}.
\end{align}
Taking the logarithm transforms a multiplicative random walk into an additive random walk.
If we now take the limit of large $n$, the law of large numbers gives
\begin{align}
\label{eq-limlogGRW}
\lim_{n\to\infty} \frac{1}{n} \left( \log X_n  - \log X_0\right) = \mathbb{E} \left[ \log z \right]
\end{align}
almost surely.
Therefore, as $n$ gets large, a typical realisation of Eq.~(\ref{eq-GRW}) almost surely behaves like
\begin{align}
\label{eq-GRW-typical}
X_n \sim X_0\, \mathrm{e}^{\lambda\,n}
\end{align}
where $\lambda =\mathbb{E} \left[ \log z \right]$ is often referred to as the Lyapunov exponent (since the typical difference between two different initial conditions $X_0^{(1)}$ and $X_0^{(2)}$ grows as 
$\left| X_0^{(1)} - X_0^{(2)} \right|\, e^{\lambda \, n}$.
On the other hand, since the $z_i$ are iid, the expected value of $X_n$ is given by
\begin{align}
\mathbb{E}\left[ X_n \right] & = X_0 \left(\mathbb{E}\left[ z \right] \right)^n.
\end{align}
Therefore as $n$ gets large
\begin{align}
\mathbb{E}\left[ X_n \right] \sim X_0 \, \mathrm{e}^{\mu\, n},
\end{align}
where $\mu = \log \mathbb{E}\left[ z \right]$.
Since the logarithm is convex, Jensen's inequality tells us that $\lambda < \mu$ in general. 
Typical trajectories are therefore not representative of the ensemble in the sense that the expected value grows exponentially faster with $n$ than individual trajectories.
The reason is that the expected value is dominated by large fluctuations which become exponentially rare as $n$ grows but make exponentially large contributions to the average. 
This mechanism is not a peculiarity of the multiplicative random walk. 
It appears in many areas of statistical physics such as localisation phenomena, glass dynamics, turbulence and random matrix products.
The multiplicative random walk is just the simplest exactly solvable model of a phenomenon that has wide applicability.

Returning to the discussion of ergodicity, the difference between individual trajectories and the ensemble averaged trajectory of the multiplicative random walk is not directly interpretable as a difference of time and ensemble averages.
 The time average of $X_n$ is either zero or infinite (depending on the sign of $\lambda$.) The ensemble average is $n$ dependent and also becomes either zero or infinite as $n \to \infty$ (depending on the sign of $\mu$.)
However, there {\em is} an underlying Birkhoff style ergodicity statement encoded in Eq.~(\ref{eq-limlogGRW}).
To see this, define a new variable, 
\begin{align}
\lambda_n = \log X_{n+1} - \log X_{n},
\end{align}
corresponding to the increment of $\log X_n$, or equivalently the local in time growth rate of $X_n$.
From Eq.~(\ref{eq-GRW}) and the fact that the $z_i$ are iid, we note that $\mathbb{E}\left[ \lambda_n \right]$ is independent of $n$ and given by $\mathbb{E}\left[ \lambda \right] = \mathbb{E}\left[ \log z \right]$.
The left hand side of Eq.~(\ref{eq-limlogGRW}) is a telescopic sum: $\log X_n - \log X_0 = \log X_n - \log X_{n-1} + \log X_{n-1} - \ldots + \log X_1 - \log X_0$.
Putting this together, Eq.~(\ref{eq-limlogGRW}) tells us that
\begin{align}
\label{eq-Birkhoff-lambda}
\lim_{n \to \infty} \frac{1}{n} \sum_{i=0}^{n-1} \lambda_i  &= \mathbb{E}\left[ \lambda \right]. 
\end{align}
This is Birkhoff's theorem, but for the growth rate of $X_n$ rather than for $X_n$ itself. 
Note what has happened here: although $X_n$ is trivially non-ergodic, by identifying a transformation of $X_n$ which does possess the ergodicity property  as per Eq.~(\ref{eq-Birkhoff-lambda}), we are able to understand the temporal behaviour of typical realisations of $X_n$ as per Eq.~(\ref{eq-GRW-typical}).

Empirical analogues of this theoretical procedure abound in practical time series analysis and signal processing. 
Transformations are commonly used to attempt to isolate a component of the dynamics that is stationary or ergodic (or approximately so in empirical contexts) for observables of interest.
The objective is usually to find transformations which can be used to improve some a-priori measure of predictive power.  
In applications, detrending and/or seasonable adjustment may be applied to remove systematic drift or periodic oscillation, differencing may be applied to remove stochastic drift and data transformations may be used to stabilise variance over time.
A very commonly used transformation for variance stabilisation is the Box-Cox transformation (which includes the log transformation discussed above as a special case):
\begin{align}
f_\eta(x) &= 
\left\{
\begin{array}{ll}
\frac{x^{1-\eta} -1}{1- \eta} & \text{if $\eta \neq 1$} \\
 \log x &  \text{if $\eta = 1$} 
\end{array}
\right.
\end{align}
and its generalisations. 
Detrending, seasonal adjustment, differencing and data transformation are often combined to find quantities that best achieve this end. 
This is reflected in the exactly solvable example of the multiplicative random walk where differencing of the log of the signal is required to obtain the quantity, $\lambda_n$, that characterises the typical behaviour of individual realisations of the process.

\section{Practical considerations}

In rigorous ergodic theory, ergodicity requires answering a question about the global invariant structure of an entire dynamical system $(\Omega, \mu, T)$:  do all invariant sets have measure 0 or 1?
In applications, researchers ask a simpler question: are empirical estimators computed from a single trajectory representative of the ensemble of possible trajectories? 
As this notes have shown, these questions are closely related theoretically.
However it is often possible to answer the latter approximately without knowing much about the former.

In practice, only a finite number of trajectories are observed and each is only of finite length, only a limited number of observables are measured and the true underlying dynamics may not even be known. 
Under such circumstances it is pretty hopeless to attempt to establish indecomposability of an invariant measure.
For this reason, weaker or approximate notions of ergodicity are sometimes used in applied work:
\begin{itemize}
\item ergodicity of the mean: does the empirical time average of a quantity of interest equal to it's empirical mean?
\item ergodicity of higher moments (or of distribution)
\item ergodicity as an approximate (FAPP etc) or  non-binary property (ergodicity spectrum etc)
\end{itemize}
These weaker notions of ergodicity provide useful ways to frame important questions in different domains of application. 
They are distinct, and more operational in nature, than the elegant definition of measure-theoretic ergodicity that has ultimately allowed mathematicians to unify the study of invariant structures in deterministic dynamical systems, stochastic processes and beyond.
However this should not preclude their use and more widespread adoption among applied scientists.
The question of when the statistics of individual trajectories are representative of populations and vice versa is a critical one in many fields. 
The concept of ergodicity too important to be solely an abstraction of mathematics and theoretical physics.

\bibliographystyle{plain}
\bibliography{refs}


\end{multicols}
\end{document}
